Global land cover fraction mapping: methods, variables, challenges

### 2020-04-28 article/article.tex
Changes from papeeria

% Metadata
\title{Fractional land cover classification method assessment using PROBA-V satellite data}
\author{Dainius Masiliūnas, Nandin-Erdene Tsendbazar, Martin Herold, \\ Myroslava Lesiv, Jan Verbesselt}
% * <Dainius Masiliūnas> 11:20:08 26 Mar 2019 UTC+0100:
% TODO: fix newline issue

@@

% Acronyms - cite with \ac{}
\newacronym{GNSS}{GNSS}{Global Navigation Satellite System}
\newacronym{OA}{OA}{overall accuracy}
\newacronym{UA}{UA}{user accuracy}
\newacronym{PA}{PA}{producer accuracy}
\newacronym{RMSE}{RMSE}{root mean squared error}
\newacronym{MAE}{MAE}{mean absolute error}
\newacronym{ME}{ME}{mean error}
\newacronym{NSE}{NSE}{Nash–Sutcliffe model efficiency coefficient}
\newacronym{SDG}{SDG}{Sustainable Development Goal}
\newacronym{LC-CCI}{LC-CCI}{ESA Climate Change Initiative Land Cover}
\newacronym{CGLS}{CGLS-LC100}{Copernicus Global Land Services Land Cover}
\newacronym{FROM-GLC}{FROM-GLC10}{Finer Resolution Observation and Monitoring Global Land Cover}
\newacronym{SMA}{SMA}{spectral mixture analysis}
\newacronym{SVM}{SVM}{support vector machine}
\newacronym{MLP}{MLP}{multi-layer perceptron}
\newacronym{NN}{NN}{neural network}
\newacronym{LOESS}{LOESS}{locally estimated scatterplot smoothing}
\newacronym[firstplural={vegetation indices (VIs)}]{VI}{VI}{vegetation index}
\newacronym{GLSDEM}{GLSDEM}{Global Land Survey Digital Elevation Model}
\newacronym{TPI}{TPI}{Terrain Position Index}
\newacronym[text={random forest}]{RF}{RF}{random forest}
\makenoidxglossaries
% Disable abbreviating RF for now
\glsunset{RF}

@@

\abstract{
\comment[id=dm]{The abstract is still a placeholder}The current common practice for producing global land cover maps is to represent land cover classes as discrete units, e.g. a pixel may be labelled as grassland, tree cover, or open water. However, in reality, the area covered by the pixel is hardly ever homogeneous, it may be e.g. trees surrounded by grass next to water. This mismatch between the reality and its representation in maps causes a loss in precision, as only the dominant land cover class can be represented in such ``hard'' classification maps. This problem is worse for coarse resolution land cover maps. A common workaround to this issue is to define mosaic classes, i.e. land cover classes that are combinations of other land cover classes, such as open forest (combination of grassland and tree cover). However, this approach leads to increasingly complex map legends and confusion about the definitions.

Our proposed solution to this issue is to employ fractional land cover mapping: for each pixel, define the proportion of the pixel that each considered land cover class occupies. This allows expressing an open forest as e.g. 40\% grassland, 60\% tree cover, and 0\% of the rest of the land cover classes. This ``fuzzy'' classification not only allows to more precisely express the land cover that exists in reality, but also gives users the flexibility to derive their own maps with a customised legend. For instance, if a user defines open forest as one that has less than 50\% tree cover and at least 30\% of grassland, they can generate a map with this definition from land cover fraction data, rather than relying on the definition that the mapmakers provide.
Exhaustive land cover fraction data is desired by a number of user communities, but to date there is no single global nor continental scale map that would provide fractions of each common land cover class.

In this paper, we tested the application of fractional land cover mapping over the continent of Africa. We compared the performance of four very different machine learning algorithms: random forest regression, multilayer perceptron neural networks, partial least squares regression, fuzzy nearest prototype classification and logistic regression. In order to allow the algorithms to make good predictions about land cover fractions, six types of covariates were used: spectral bands and vegetation indices derived from the Proba-V 100 m TOC reflectance data; temporal features derived from the full time series of said imagery; terrestrial features from a digital elevation model; soil properties and climate data from global models, and location. We compared the contribution of each of the covariate types to improving the accuracy of the model predictions.

@@

The models were trained on 26668 training points throughout Africa. The training points, collected as part of the Copernicus Global Land System: Dynamic Land Cover (CGLS-LC100) project, describe the fraction that each land cover class (bare soil, crops, grassland, shrubs, trees, urban/built-up, water and wetlands) occupies within a 100\texttimes100 m area, which is aligned to the grid used for Proba-V 100 m products. After the models were trained, their land cover class fraction predictions were compared with an independent validation set of 3617 points, which are likewise aligned to the Proba-V grid. Overall as well as per-class RMSE, MAE, and ME statistics were used to compare the models. In addition, the subpixel confusion-uncertainty matrix was used to analyse which classes were the most difficult to separate for each model, as well as to derive overall, users' and producers' accuracies and the kappa statistic. The models were compared with each other, as well as against a control intercept-only model.

Preliminary results show that the three-model approach increases the overall accuracy of the predictions by more accurately modelling the extreme values. From the machine learning models that were compared, random forest regression reaches the highest overall accuracy of 71\%. This is comparable to the reported accuracies of the current ``hard'' global land cover maps. Covariate importance analysis shows that spectral covariates are the most important for predicting land cover fractions, followed by the location, terrestrial and temporal covariates.

The application of the method on data gathered throughout Africa shows the applicability of the method to a variety of land cover conditions, ranging from highly heterogeneous agricultural landscapes to homogeneous bodies of water. Given enough reference data from other parts of the world and enough computational power, the method could be applied globally to upscale it to a global fractional land cover map. Such a map would cater to users with different fields of focus. Among other benefits, more precise estimation of land cover allows for 1) improved climate models to help reach Sustainable Development Goals (for which land cover is a key variable), 2) facilitate the making of models for vegetation and urban area dynamics, and 3) help governments and land owners keep better track of relevant land cover and its change. In addition, this work contributes to the operationalisation of the CGLS-LC100 project, which aims to produce exhaustive global land cover fractions as an operational product.

@@

\section{Introduction}
% * <Dainius Masiliūnas> 14:50:56 29 Jul 2019 UTC+0200:
% ~1250 words, 8 paragraphs + objectives

@@

Land cover, as one of the key variables for monitoring a number of \cglspl{SDG}, has received attention in recent years, and new global land cover maps have emerged to better map the current land cover of the world, as well as to track land cover change.
The most recent achievements have been the \ac{LC-CCI} map \highlight[id=dm, comment=Is there a better citation?]{\citep{defourny2012cci}} that aims at maintaining a long medium-resolution time series of global land cover for the climate community, \highlight[id=dm, comment=ditto]{\ac{CGLS}} map that aims at a finer spatial resolution and high quality with yearly updates since 2015, and the \ac{FROM-GLC} map \citep{fromglc2019} that showcases the potential of land cover mapping at 10 m resolution.

@@

A potential solution to this issue is land cover fraction mapping, where instead of a discrete class, the proportion of every class in the legend is reported for every pixel of the map.
This is also called ``fuzzy'' or ``soft'' classification, and sometimes ``subpixel'' classification or ``linear mixture modelling'' \citep{Okeke2006fuzzyexponent}. A related term is ``super
Land cover fraction mapping has been attempted in the past, both in research \citep{adams_classification_1995, foody1996fuzzyevaluation, colditz_land_2011, sharma_assessing_2011, uma_shankar_wavelet-fuzzy_2011, dwivedi_optimisation_2012, lizarazo_quantitative_2012, gessner_estimating_2013, okujeni_generalizing_2018} and production (e.g. \citealp{hansen2000hardtree, hansen_continuous_2011, pengra_global_2015, hansen_global_2003, sexton_global_2013}).
Nevertheless, to date there is \highlight[id=dm, comment=Depends on how we position CGLOPS]{no global map} that would provide fractions of every major land cover class in a single product.

Land cover fraction mapping can be performed using different approaches and algorithms.
In its core, it is a regression rather than a classification problem, as the output is a fraction of a label rather than a label itself.
Popular approaches include fuzzy nearest centroid regression \citep{zhang2001fullyfuzzy}, \ac{SMA} (e.g. \citealp{yang_landsat_2012, hobbs2003linear, shimabukuro1991least}), \ac{RF} regression \citep{schwieder_estimating_2014}, \gls{SVM} regression \citep{schwieder_estimating_2014}, \gls{MLP} \glspl{NN} \citep{zhang2001fullyfuzzy}, genetic algorithms \citep{stavrakoudis_boosted_2011} and wavelet transformation \citep{uma_shankar_wavelet-fuzzy_2011}.

@@

In addition to classification methods, one aspect that is important for classification accuracy is the data (covariates) that the algorithm is trained on and predict land cover from \citep{yu2014metadiscoveries}. In addition to spectral covariates (blue, red, near infrared, shortwave infrared bands), temporal covariates (growing season start, duration, growing intensity) are commonly used to improve the separability of crops \citep{jakubauskas2001harmonic}. Elevation covariates (elevation, slope, aspect) are known to improve the separability between tree classes \citep{burrough2001fuzzy}. \Glspl{VI} have been used to improve the detection of wetlands \citep{sader1995wetlands}, as well as for defining classification rules in rule-based fuzzy classification \citep{baraldi2006rulebased}. Therefore in this thesis such covariates have also been used in order to increase land cover classification accuracy, and their importance for prediction was analysed.

\textbf{Objectives}:

\begin{enumerate}
 \item Compare the performance of a variety of machine learning models for a global-scale land cover fraction mapping task
 \item Compare the importance of covariates and covariate groups for the prediction of land cover fractions
 \item Investigate methods for reducing bias in the predictions with regards to zero inflation and predictions tending towards the mean
\end{enumerate}

@@

For training the global land cover fraction models, a training dataset collected by IIASA was used. This global point dataset provides information on nine class fractions of sampled pixels of the Proba-V 100 m product.
The fraction estimate was acquired by high-resolution image interpretation by regional experts, for the year 2015.
Each of the sampled Proba-V pixels was subdivided into a 10-by-10 subpixel grid and the area estimates were derived from the subpixel ratios.

For model validation, an independent global dataset collected by Wageningen University \& Research was used.
The method of point collection is equivalent to the method used for training data collection, but performed by separate teams of regional experts.

The data includes over 150\,000 training points and over 20\,000 validation points across the globe (see figure \ref{fig-reference-data}).
Since the data describes fractions, it is zero-inflated, since it is relatively rare for more than just a few land cover classes to occupy a single pixel; the rest are set to zero.

\begin{figure}
 \includegraphics[width=\textwidth]{article-figures/maps/2019-08-06-training-and-validation}
 \caption{Reference data points with the dominant land cover class. (a): training dataset. (b): reference dataset.}
 \label{fig-reference-data}
\end{figure}

@@

To predict land cover fractions in unsampled areas, the models learn from covariates present at the locations of the training data, and then the covariates are used for prediction in unsampled areas.
In this study, five groups of covariates were used: optical, terrain, soil, climate and location.
All of these covariates had to be preprocessed before they could be input into the models.

%\begin{enumerate}
% \item GLSDEM
% \item Proba-V time series
% \item SoilGrids and LandGIS
% \item WorldClim
%\end{enumerate}

@@

The entire archive of the Proba-V 100 m Level 3 Top-of-Canopy 5-day composite product was used for this study.
The time series of NDVI and each of the bands at the points of interest were cloud masked, first by applying the status mask provided with the product itself, and then by running a temporal cloud filter to remove the remaining outliers that were too far from the fitted \ac{LOESS} curve over the blue band.

@@

Lastly, extra \glspl{VI} (NDMI, EVI, OSAVI and NIR\textsubscript{v}) were calculated.
Each of these \glspl{VI} was then composited into a single image that represents the leaf-on season of 2015.

@@

The NASA \ac{GLSDEM} 90 m product was used as the elevation covariate.
In addition, terrain parameters were calculated out of elevation: slope, aspect, roughness and \ac{TPI}.
All terrain covariates were then resampled to the Proba-V grid.

@@

The climate data from WorldClim 2.0 30 s product was used.
Biophysical parameters were calculated from the data as well.
Some additional biophysical parameters were also calculated, namely the means, minimums and maximums over the whole time series of solar radiation, wind speed and water vapour pressure.

@@

\subsubsection{Location covariates}

The latitude, longitude and absolute latitude of the points were also included as covariates.

@@

\begin{enumerate}
 \item Temporal cloud filter for Proba-V
 \item Harmonic metrics extraction
 \item Terrain metrics calculation
 \item Additional climate parameter calculation
 \item Manual model feature selection
\end{enumerate}

@@

\begin{figure}
 \includegraphics[width=\textwidth]{article-figures/flowcharts/processing}
 \caption{Processing chain.}
 \label{fig-processing}
\end{figure}

\subsection{Fractional land cover mapping methods}

\subsubsection{Logistic regression}

@@

\subsubsection{Partial least squares regression}

\subsubsection{Fuzzy nearest prototype/centroid}

\subsubsection{Neural networks}

AKA multilayer perceptron

\subsubsection{Random forest regression}

Including variable importance

@@

As the dataset is unbalanced (dominated by zeroes and hundreds) and thus the objective function tends to draw the model towards good prediction of 0/100, whereas in regression the middle is also important. Therefore we tried several approaches to increase the accuracy of the middle predictions.

@@

Not sure if this should be a section, or just part of results:

\begin{enumerate}
 \item \ac{RMSE}, \ac{MAE}, \ac{ME} per class
 \item Sub-pixel confusion-uncertainty matrix: kappa, \ac{OA}, \ac{PA}, \ac{UA}
 \item Correlation matrix
 \item Prediction accuracy at different parts of the predicted vs observed line
 \item Comparison with a control/intercept model (10\% of everything)?
\end{enumerate}

>>>>

% Metadata
\title{Global land cover fraction mapping: methods, variables, challenges}
\author{Dainius Masiliūnas, Nandin-Erdene Tsendbazar, Martin Herold, \\ Myroslava Lesiv, Marcel Buchhorn, Jan Verbesselt}
% * <Dainius Masiliūnas> 11:20:08 26 Mar 2019 UTC+0100:
% TODO: fix newline issue

@@

% Acronyms - cite with \ac{}
\newacronym{GNSS}{GNSS}{Global Navigation Satellite System}
\newacronym{OA}{OA}{overall accuracy}
\newacronym{UA}{UA}{user accuracy}
\newacronym{PA}{PA}{producer accuracy}
\newacronym{RMSE}{RMSE}{root mean squared error}
\newacronym{MAE}{MAE}{mean absolute error}
\newacronym{ME}{ME}{mean error}
\newacronym{NSE}{NSE}{Nash–Sutcliffe model efficiency coefficient}
\newacronym{SDG}{SDG}{Sustainable Development Goal}
\newacronym{LC-CCI}{LC-CCI}{ESA Climate Change Initiative Land Cover}
\newacronym{CGLS-LC100}{CGLS-LC100}{Copernicus Global Land Services 100 m Land Cover}
\newacronym{CGLS}{CGLS}{Copernicus Global Land Services}
\newacronym{FROM-GLC}{FROM-GLC10}{Finer Resolution Observation and Monitoring Global Land Cover}
\newacronym{IIASA}{IIASA}{International Institute for Applied Systems Analysis}
\newacronym{SMA}{SMA}{spectral mixture analysis}
\newacronym{SVM}{SVM}{support vector machine}
\newacronym{MLP}{MLP}{multi-layer perceptron}
\newacronym{NN}{NN}{neural network}
\newacronym{LOESS}{LOESS}{locally estimated scatterplot smoothing}
\newacronym[firstplural={vegetation indices (VIs)}]{VI}{VI}{vegetation index}
\newacronym{GLSDEM}{GLSDEM}{Global Land Survey Digital Elevation Model}
\newacronym{TPI}{TPI}{Terrain Position Index}
\newacronym{OSAVI}{OSAVI}{Optimised Soil-Adjusted Vegetation Index}
\newacronym{NDVI}{NDVI}{Normalised Difference Vegetation Index}
\newacronym{NIR}{NIR}{near infra-red}
\newacronym{SCM}{SCM}{subpixel confusion-uncertainty matrix}
\newacronym{FNC}{FNC}{fuzzy nearest centroid}
\newacronym{PLS}{PLS}{partial least squares}
\newacronym{MLR}{MLR}{multinomial logistic regression}
\newacronym{RF}{RF}{random forest}
\newacronym{TOC}{TOC}{Top-of-Canopy}
\newacronym{CART}{CART}{Classification and regression tree}
\makenoidxglossaries
% Disable abbreviating RF for now [text={random forest}]
%\glsunset{RF}

@@

\abstract{The current common practice for producing global land cover maps is to represent land cover classes as discrete units, e.g. a pixel may be labelled as grassland, tree cover, or open water. However, in reality, the area covered by the pixel is hardly ever homogeneous, it may be e.g. trees surrounded by grass next to water. This mismatch between the reality and its representation in maps causes a loss in precision, as only the dominant land cover class can be represented in such ``hard'' classification maps. This problem is worse for coarse resolution land cover maps. A common workaround to this issue is to define mosaic classes, i.e. land cover classes that are combinations of other land cover classes, such as open forest (combination of grassland and tree cover). However, this approach leads to increasingly complex map legends and confusion about the definitions.

Our proposed solution to this issue is to employ fractional land cover mapping: for each pixel, define the proportion of the pixel that each considered land cover class occupies. This allows expressing an open forest as e.g. 40\% grassland, 60\% tree cover, and 0\% of the rest of the land cover classes. This ``fuzzy'' classification not only allows to more precisely express the land cover that exists in reality, but also gives users the flexibility to derive their own maps with a customised legend. For instance, if a user defines open forest as one that has less than 50\% tree cover and at least 30\% of grassland, they can generate a map with this definition from land cover fraction data, rather than relying on the definition that the mapmakers provide.
Thus exhaustive land cover fraction data is desired by a number of user communities.

In this paper, we compared the performance of eight very machine learning algorithms for land cover fraction mapping: random forest regression, multilayer perceptron neural networks, partial least squares regression, fuzzy nearest prototype classification, multinomial logistic regression, lasso regression, cubist regression and support vector regression. In order to allow the algorithms to make good predictions about land cover fractions, six types of covariates were used: spectral bands and vegetation indices derived from the Proba-V 100 m TOC reflectance data; temporal features derived from the full time series of said imagery; terrestrial features from a digital elevation model; soil properties and climate data from global models, and location. We compared the contribution of each of the covariate types to improving the accuracy of the model predictions.

@@

The models were trained on 138164 training points throughout the globe. The training points, collected as part of the Copernicus Global Land System: Dynamic Land Cover (CGLS-LC100) project, describe the fraction that each land cover class (bare soil, crops, grassland, shrubs, trees, urban/built-up, water and wetlands) occupies within a 100\texttimes100 m area, which is aligned to the grid used for Proba-V 100 m products. After the models were trained, their land cover class fraction predictions were compared with an independent validation set of 20705 points, which are likewise aligned to the Proba-V grid. Overall as well as per-class RMSE, MAE, and ME statistics were used to compare the models. In addition, the subpixel confusion-uncertainty matrix was used to analyse which classes were the most difficult to separate for each model, as well as to derive overall, users' and producers' accuracies and the kappa statistic. The models were compared with each other, as well as against a control intercept-only model.

Results show that the three-model approach increases the overall accuracy of the predictions by more accurately modelling the extreme values. From the machine learning models that were compared, random forest and cubist regressions reaches the highest overall accuracy of 72\%. This is comparable to the reported accuracies of the current ``hard'' global land cover maps. Random Forest covariate importance analysis shows that spectral covariates are the most important for predicting land cover fractions, followed by the location, terrestrial and temporal covariates.

The application of the method on data gathered throughout the globe shows the applicability of the method to a variety of land cover conditions, ranging from highly heterogeneous agricultural landscapes to homogeneous bodies of water. This research proves that machine learning algorithms can be applied globally to create a global fractional land cover product. Such a product would cater to users with different fields of focus. Among other benefits, more precise estimation of land cover allows for 1) improved climate models to help reach Sustainable Development Goals (for which land cover is a key variable), 2) facilitate the making of models for vegetation and urban area dynamics, and 3) help governments and land owners keep better track of relevant land cover and its change. In addition, this work contributes to the operationalisation of the CGLS-LC100 project, which aims to produce exhaustive global land cover fractions as an operational product.

@@

\pagebreak
\tableofcontents
% * <Dainius Masiliūnas> 11:40:16 21 Apr 2020 UTC+0200:
% Temporary, to make it easier to understand the structure
\pagebreak

\section{Introduction}

@@

Land cover, as one of the key variables for monitoring a number of \glspl{SDG}, has received attention in recent years, and new global land cover maps have emerged to better map the current land cover of the world, as well as to track land cover change.
Some of the recent achievements have been the \ac{LC-CCI} map \highlight[id=dm, comment=Is there a better citation?]{\citep{defourny2012cci}} that aims at maintaining a long medium-resolution time series of global land cover for the climate community, \highlight[id=dm, comment=ditto]{\ac{CGLS-LC100}} map that aims at a finer spatial resolution and high quality with yearly updates since 2015, and the \ac{FROM-GLC} map \citep{fromglc2019} that showcases the potential of land cover mapping at 10 m resolution.

@@

A potential solution to this issue is land cover fraction mapping, where instead of a discrete class, the proportion of every class in the legend is reported for every pixel of the map.
This is also called ``fuzzy'' or ``soft'' classification, and sometimes ``subpixel'' classification or ``linear mixture modelling'' \citep{Okeke2006fuzzyexponent}. A related term is ``super
Land cover fraction mapping has been attempted in the past, both in research \citep{adams_classification_1995, foody1996fuzzyevaluation, colditz_land_2011, sharma_assessing_2011, uma_shankar_wavelet-fuzzy_2011, dwivedi_optimisation_2012, lizarazo_quantitative_2012, gessner_estimating_2013, okujeni_generalizing_2018} and production (e.g. \citealp{hansen2000hardtree, hansen_continuous_2011, pengra_global_2015, hansen_global_2003, sexton_global_2013}).
Nevertheless, to date only \ac{CGLS-LC100} provides a global map with fractions of every major land cover class.

Land cover fraction mapping can be performed using a variety of different approaches and algorithms.
In its core, it is a regression rather than a classification problem, as the output is a fraction of a label rather than a label itself.
Popular approaches include fuzzy nearest centroid regression \citep{zhang2001fullyfuzzy}, \ac{SMA} (e.g. \citealp{yang_landsat_2012, hobbs2003linear, shimabukuro1991least}), \ac{RF} regression \citep{schwieder_estimating_2014}, \gls{SVM} regression \citep{schwieder_estimating_2014}, \gls{MLP} \glspl{NN} \citep{zhang2001fullyfuzzy}, genetic algorithms \citep{stavrakoudis_boosted_2011} and wavelet transformation \citep{uma_shankar_wavelet-fuzzy_2011}.
So far other studies have only compared a few of these methods at once, and never with an exhaustive set of land cover classes nor a global scale.

@@

In addition to classification methods, one aspect that is important for classification accuracy is the data (covariates) that the algorithm is trained on and predict land cover from \citep{yu2014metadiscoveries}. In addition to spectral covariates (blue, red, near infrared, shortwave infrared bands), temporal covariates (growing season start, duration, growing intensity) are commonly used to improve the separability of crops \citep{jakubauskas2001harmonic}. Elevation covariates (elevation, slope, aspect) are known to improve the separability between tree classes \citep{burrough2001fuzzy}. \Glspl{VI} have been used to improve the detection of wetlands \citep{sader1995wetlands}, as well as for defining classification rules in rule-based fuzzy classification \citep{baraldi2006rulebased}. Therefore in this paper such covariates have also been used in order to increase land cover classification accuracy, and their importance for prediction was analysed.

All in all, with this study we aim to investigate and compare methods for performing land cover fraction mapping, at a global scale, and with a comprehensive set of methods as well as a wide variety of input covariates, so as to improve land cover fraction map production.
We also aim to suggest ways of dealing with the balance issues inherent to land cover fraction reference data that hamper land cover fraction mapping.
Therefore the objectives of our study are:

\begin{enumerate}
 \item Compare the performance of a variety of machine learning models for a global-scale land cover fraction mapping task
 \item Investigate approaches for reducing bias in the predictions with regards to zero inflation and predictions tending towards the mean
 \item Compare the importance of covariates and covariate groups for the prediction of land cover fractions
\end{enumerate}

@@

The reference data (for model training and validation) used in this study was collected as part of the \ac{CGLS} project.
The training set was generated by experts in \ac{IIASA}.
This global point dataset provides information on nine class fractions of sampled pixels of the Proba-V 100 m product.
The fraction estimate was acquired by high-resolution image interpretation by experts, for the year 2015.
Each of the sampled Proba-V pixels was subdivided into a 10-by-10 subpixel grid and the area estimates were derived from the subpixel ratios.

For model validation, an independent global dataset of \ac{CGLS}, generated by Wageningen University \& Research \citep{tsendbazar_developing_2018}, was used.
The method of point collection is equivalent to the method used for training data collection, but performed independently by a different group of regional experts.

The data includes over 150\,000 training points and over 20\,000 validation points across the globe, describing the fractions of 12 classes (see figure \ref{fig-reference-data}) as they were in the year 2015.
Since the data describes fractions, it is zero-inflated, since it is relatively rare for more than just a few land cover classes to occupy a single pixel.
This means that for a given pixel, only a few of the 12 classes would normally have a fraction that is above 0\%.
This leads to a dataset that is zero-inflated (also known as sparse), with over half of all values for a pixel being zero.
Such an imbalance is a problem for statistical and machine learning techniques that are tuned by minimising the error of the output, since they are heavily biased towards a good prediction of zeroes.
In other words, when validating against a zero-inflated dataset, a prediction of zero is a safe bet.
But for user communities, correct prediction of non-zero values is more important.

\begin{figure}
 \includegraphics[width=\textwidth]{article-figures/maps/2019-08-06-training-and-validation}
 \caption{Reference data points with the dominant land cover class. (a): training dataset. (b): validation dataset.}
 \label{fig-reference-data}
\end{figure}

@@

To predict land cover fractions in unsampled areas, the models learn from covariates present at the locations of the training data, and then the covariates are used for prediction in unsampled areas.
In this study, six groups of covariates were used: spectral, harmonic, terrain, soil, climate and location (see table \ref{tab-inputdata}).
All of these covariates had to be preprocessed before they could be input into the models.

\begin{table}
\centering
    \begin{tabular}{llp{6cm}}
         \toprule
         \textbf{Category} & \textbf{Number} & \textbf{Data source} \\
         \midrule
         Spectral & 22 & Proba-V 100m \ac{TOC} reflectance v1.02 \citep{probavguide2} \\
         Harmonic & 9 & Proba-V 100m \ac{TOC} reflectance v1.02 \citep{probavguide2} \\
         Terrain & 4 & ASTER GDEM V003 \citep{ASTGTM003} \\
         Climate & 21 & WorldClim 2.0 \citep{worldclim2} \\
         Soil & 8 & SoilGrids \citep{hengl_soilgrids250m_2017} \\
         Location & 3 & None (intrinsic) \\
         \bottomrule
    \end{tabular}
    \label{tab-inputdata}
    \caption{Data sources for the covariates input into the model.}
\end{table}

@@

The entire archive (2014-03-11 to 2019-07-16) of the Proba-V 100 m Level 3 Top-of-Canopy 5-day composite product \citep{probavguide2} was used for this study.
The time series of NDVI and each of the bands at the points of interest were cloud masked, first by applying the status mask provided with the product itself, and then by running a temporal cloud filter to remove the remaining outliers that were too far from the fitted \ac{LOESS} curve over the blue band.

@@

Lastly, extra \glspl{VI} (NDMI, EVI, OSAVI and NIR\textsubscript{v}) were calculated.
Each of these \glspl{VI} was then composited into a single image that represents the leaf-on season (three months) of 2015.

@@

The ASTER GDEM v003 \citep{ASTGTM003} product was used as the elevation covariate.
In addition, terrain parameters were calculated out of elevation: slope, aspect and \ac{TPI}.
All terrain covariates were then resampled to the Proba-V grid.

@@

The climate data from WorldClim 2.0 30 s product \citep{worldclim2} was used.
Biophysical parameters were calculated from the data as well.
Some additional biophysical parameters were also calculated, namely the means, minimums and maximums over the whole time series of solar radiation, wind speed and water vapour pressure.

@@

Two products were used for data on soils: SoilGrids and LandGIS.
SoilGrids is based on a machine learning model that predicts soil properties at various soil depths globally at a 250 m resolution \citep{hengl_soilgrids250m_2017}
LandGIS is a successor to SoilGrids using a refined machine learning methodology \citep{hengl_predictive_2019}, however, it does not yet provide as many covariates as SoilGrids.

In the creation of SoilGrids, a land cover map (based on MODIS) was used.
In order to avoid circular dependencies, the covariates that are significantly influenced by the land cover map as detailed in \citet{hengl_soilgrids250m_2017} were excluded.
In addition, the taxonomy covariates were also excluded, since they are categorical data derived from the numerical soil property data.

\subsubsection{Other covariates}

The latitude, longitude and absolute latitude of the points were also included as location covariates.
Lastly, as part of the SoilGrids dataset, the MODIS land surface temperature product (MOD11A2, \citealp{wan_mod11a2_2015}) data monthly composites for 2011 were also used as additional covariates.

@@

In total, 313 covariates were used for predicting the land cover fractions.
However, many of them are colinear with one another, and thus variable selection was used to remove colinear covariates.
Covariates that had a Pearson's correlation coefficient $r$ above 0.9 were excluded in an iterative process.
After that, covariates with a Spearman's $\rho$ above 0.9 were likewise excluded.

Most of the colinear covariates were soil covariates at different depths.
Therefore the 10 cm depth covariates were left in, and the other depths excluded, as long as $r$ was above 0.9.
Similarly, climate data was colinear between subsequent months. Thus January and August data was preferred, as these months represent different extremes of the year and are less correlated with other months' data.

While \ac{OSAVI} was calculated as a covariate, ultimately it was dropped due to too high correlation with \ac{NDVI}.
Similarly, the spectral bands of Proba-V were also highly correlated, so only the red and \ac{NIR} bands were kept.

After the covariate selection process, 67 covariates remained.
These covariates include data from each of the covariate categories.
See table \ref{tab-inputdata} for an overview.

@@

\begin{figure}
 \includegraphics[width=\textwidth]{article-figures/flowcharts/processing}
% * <Dainius Masiliūnas> 13:45:26 13 Aug 2019 UTC+0200:
% TODO: optimise style for readability.
 \caption{Processing chain.}
 \label{fig-processing}
\end{figure}

\subsection{Land cover fraction mapping methods}

For deriving land cover fractions, we compared a wide array of machine learning regression methods: \ac{FNC}, \ac{MLP} \glspl{NN}, \ac{PLS} regression, Lasso regression, \ac{MLR}, \ac{RF} regression.
% * <Dainius Masiliūnas> 16:05:16 07 Aug 2019 UTC+0200:
% Perhaps also SVM and MESMA
In addition, as a baseline we also compare the results with an intercept model (all fractions always predicted to be equal).

All of the models were made using the free and open-source statistical software R \citep{r_2019}.

@@

Ridge regression is a regularisation method which reduces large coefficients of model covariates in order to prevent overfitting.
Lasso regression is based on the same principle, except it reduces the coefficients of each model covariate to zero if the absolute sum of coefficients is above a threshold, thus acting as a variable selection method.
Elastic net regression is a method combining the two, where model coefficients may be reduced or set to zero.
This has the effect of the coefficients of correlated covariates being set to a similar value.

Since we performed a manual covariate selection prior to running regularisation to remove highly correlated covariates, and having a further reduced set of covariates is useful for other algorithms, for this method we chose Lasso regression.
Lasso regression builds upon least squares, and thus the classes are predicted separately to each other, using the binary relevance approach as presented by \citet{karalas2016br}.
Therefore, the output fractions do not necessarily sum to 100\%.
As such, they were linearly scaled.
The lasso regression was performed using the \texttt{glmnet} package \citep{glmnet}.

\subsubsection{\Glsentryfull{PLS} regression}

\ac{PLS} regression is a concept related to principal component regression and also aims at reducing the input covariates into latent variables that explain the most variance of the response variable in multidimensional space \citep{martens_multivariate_1992}.
As such, it is well suited for multicollinear data.
Due to this property, the \highlight[id=nt,comment={would you not agree for better comparability just use the same set of covariates?}]{full set of covariates} were used as input to this model, rather than the reduced (uncorrelated) set.
\ac{PLS} regression was performed using the package \texttt{pls} \citep{pls}.

\subsubsection{\Glsentryfull{FNC} regression}

\Gls{FNC} regression, also called fuzzy nearest prototype, fuzzy $c$-means or fuzzy $k$-means, is a simple regression method where the pixel's membership of a class is determined by the distance of the pixel from the class centroid in feature space.
This method always produces output that sums to 100\%.
In this study, we used an implementation adapted from the package GSIF \citep{hengl2004fuzzycmeans}.

\subsubsection{\Glsentryfull{MLR}}

While logistic regressions are most commonly used for classification tasks, the output of a multinomial regression itself is the probability of a pixel to belong to each of the defined classes.
This probability can be used as a proxy for the fraction of the pixel that the class covers.
These probabilities always add up to 100\%.
However, an \ac{MLR} can only be trained on labelled data, rather than fractions.
Therefore the target (label) input for the \ac{MLR} was the name of the dominant land cover fraction of each pixel.
This does not guarantee that the pixels that the model was trained on were pure endmembers, but it allows the model to learn from a much larger dataset compared to using just the pure pixels.
The logistic regression implementation was provided by the \texttt{nnet} package \citep{nnet}.

\subsubsection{\Glsentryfullpl{NN}}

\Glspl{NN} are a promising technique for land cover fraction mapping, as they allow both multiple inputs and multiple outputs, and, using the softmax activation function, also ensures that the result sums up to 100\%.
In this study, we used a multilayer perceptron with two layers of 256 neurons and softmax activation.
The \texttt{keras} package \citep{keras}, built upon TensorFlow, was used as the \ac{NN} implementation, and for the ability to make use of a graphics card to accelerate the neural network training process.

\subsubsection{\Glsentryfull{RF} regression}

\ac{RF} regression is a popular method for land cover classification that works by building a number of decision trees based on random subsets of the input training data, and averaging the ``votes'' of these individual trees.
\ac{RF} can only take one variable as the response variable, therefore there are two ways to obtain land cover fraction predictions from an \ac{RF} model.
One way is analogous to how we performed \ac{MLR}: giving labels of the dominant land cover class as the response variable, and then considering the proportion of votes from the trees as the proportions of land cover classes in the pixel.
The second way is binary relevance, where a separate model is trained to predict the presence of each class.
Then the predictions are scaled to sum up to 100\%.
In this study, we chose the second approach, so that the model for each class would be able to learn from the fractions of all the input data, rather than just the labels of the dominant land cover class.
The processing was done using the \texttt{ranger} package \citep{ranger}.

\ac{RF} was also used for assessing the covariate importance.
Since the number of covariates is imbalanced by group (i.e. there are merely 3 location covariates and 168 soil covariates, which are often collinear), covariate importance was assessed by permuting all covariates in a group and reporting the decrease in \ac{RMSE} resulting from that.

@@

After determining the method that performs the best, we attempted to tune it further in an attempt to solve the dataset balance (zero inflation) problem.
As the dataset is dominated by zeroes and hundreds, minimising the objective function tends to draw the model towards predicting 0\% fractions well, and ignoring the prediction errors in the middle of the 0-100\% range.
This is not desirable for users of land cover fraction data, as the added value of fractions is the information about the middle of the range; otherwise, discrete classification would be just as good.
Therefore we tried several approaches to increase the accuracy of the middle predictions.

@@

To assess the performance of the models, we used a number of statistical measures:
